---
title: Test Post
author: Andrew Kostandy
date: '2018-03-28'
slug: test-post
categories: []
tags:
  - r
  - python
---

This is a test post.  

Some random text: This will show up in the summary under the post title. This post was written in R Markdown. It shows that the functionality of using R and Python code in R Markdown works well. The output is rendering successfully and the code highlighting looks good. It's also really easy to update a post and the process is seamless.

Here's a random equation: $y=\frac{x^2 + 55x-4}{14}$  

# R Code

```{r, message=FALSE}
library(tidyverse)
```

```{r}
ggplot(mtcars, aes(mpg,hp)) + geom_point()
```

# Python Code

```{r, echo=FALSE}
knitr::knit_engines$set(python = reticulate::eng_python)
```

```{python, engine.path="/Users/andrewk/anaconda3/bin/python", results='hide'}
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sb
from ggplot import *
```

```{python, engine.path="/Users/andrewk/anaconda3/bin/python"}
x = np.arange(0,10)
print(x)
```

```{python, engine.path="/Users/andrewk/anaconda3/bin/python", fig.width=6, fig.height=6}
df2 = pd.DataFrame({
      'Country':pd.Categorical(["China","China","China","China","China","China",
                                "China","China","US","US","US","US","US","US",
                                "US","US"]),
      'Gender':pd.Categorical(["M","M","M","M","F","F","F","F","M","M","M","M",
                               "F","F","F","F"]), 
      'Values' : np.array([1,2,2,4,2,3,3,5,1,5,3,6,4,6,3,7])
      })
      
plt.figure()
sb.violinplot(x="Country", y="Values", hue="Gender",
              palette="muted", split=True, data=df2)
plt.show()
```